Ensemble Stability of Biological Networks

### 2014-09-12 paper.tex
add todonotes and autoref labelling

% user added
\usepackage{longtable}
\usepackage{dot2texi}
\usepackage{tikz}
\usetikzlibrary{automata,shapes,arrows}

@@

@@

@@

\section{Stability analysis of biological networks}\label{sec:stabanalbn}

Consider an n-dimensional vector
$$
x(t) = (x_1(t), \ldots x_n(t)) = \vec{x}(t)
$$
whose components are solutions to the arbitrary first order system
$$
\frac{dx_i(t)}{dt} = F_i(\vec{x}(t)), \; (i=1,\ldots,n)
$$
where $F_i$ represent, potentially nonlinear, functions of the given vector of state variables. We can associate a directed graph $G_X$ that describes the manner in which each of the variables depends upon one another given by an adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}
   \right.
\end{displaymath}
For example, given the system $X = \{\{x_1,x_2\},\{F_1,F_2\}\}$
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}
we have the graph
\begin{center}
\input{tex/examplehamdyngraph}
\end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$

@@

The spectral abscissa of the matrix $A$ is defined as
$$
\eta(A) = \max_i \{\Re(\lambda_i)\}
$$
where $\lambda_i$ are the eigenvalues of $A$. The system defined by $F_i$ and $\vec{x}^0$ is dynamically stable if the spectral abscissa of $A$ is less than zero, equivalently, $\eta(A) < 0$. This is because the general solution to equation \ref{eq:lineardynsys} is
$$
y_i(t) = \sum_j b_{ij} e^{\lambda_j t}, \; (i=1,\ldots,n)
$$
for some matrix $B=(b_{ij})$ and thus all $\vec{y} = \vec{x} - \vec{x}^0$ decay to zero when all $\lambda_i < 0$.

@@

\section{Stability criteria}
Suppose that $M$ is an $n \times n$ matrix with real coefficients. Motivated by the discussion of Section \ref{sec:stabanalbn}, we will call the matrix $M$ stable if all its eigenvalues have negative real part.  Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$. Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$, writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.  Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that $M$ be stable is equivalent to asking that the roots of $\chi$ all have negative real part.  We will formulate conditions for this to happen in terms of the coefficients of $\chi$.

@@

The simplest case is $n=2$.  There we have the regions $R_{2000},
R_{1100}, R_{0200}, R_{0020}, R_{0002}$ and our invariants are $D =
s_2, \delta = s_1^2 - 4s_2, \Delta = s_1$.  The relation between these
is summarized by the conditions
\begin{align}
R_{2000} \qquad &D > 0, \delta > 0, \Delta > 0, \cr
R_{1100} \qquad &D < 0, \delta > 0, \cr
R_{0200} \qquad &D > 0, \delta > 0, \Delta < 0, \cr
R_{0020} \qquad &\delta < 0, \Delta > 0, \cr
R_{0002} \qquad &\delta < 0, \Delta < 0. \cr
\end{align}
These regions used for the classification of stability of two-component systems are show in Fig. \ref{fig:region2x2}. The stable regions are given by $R_{2000}$ and $R_{0020}$. All of the other regions represent unstable matrices.

@@

Moving to $n=3$, our invariants now look as follows: $D = s_3, \delta
= 18 s_1 s_2 s_2 + s_1^2 s_2^2 - 4 s_2^3 - 4 s_1^3 s_3 - 27 s_3,
\Delta = s_1 s_2 - s_3$.  However, now we can no longer simply use their
signs to distinguish regions.  For instance, consider the regions
$R_{3000}$ and $R_{1200}$.  These both have $D > 0$ and $\delta > 0$
but to tell them apart, we need to note that the region $\delta > 0$
has two connected components and use the auxiliary condition that $s_1 > 0$ to
specify which of the components contains our region. The regions analogous to Fig. \ref{fig:region2x2} for three component systems where $D=1$ is given in Fig. \ref{fig:region3x3}.
%(Looking at the sign of $\Delta$ will not help here because, while $\Delta > 0$ in the region $R_{3000}$, it is also positive for some elements of $R_{1200}$ as well.)

@@

The probability of \emph{a priori} stability and of stability to perturbations for all three component systems is given in Table \ref{tab:structstabmat3}. Fig. \ref{fig:connectcycle3D3x3} shows the probability of stability to perturbations for all three component systems as a function of connectivity and number of cycles. Fig. \ref{fig:stab3x3} shows the probability of stability to perturbations for all three component systems as a function of connectivity. Fig. \ref{fig:cycle3x3} shows the probability of stability to perturbations for all three component systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}.

@@

\input{tex/tablewithgraphs}

>>>>

% user added
\usepackage{longtable}
\usepackage{dot2texi}
\usepackage{tikz}
\usetikzlibrary{automata,shapes,arrows}
% todo notes see http://www.texample.net/tikz/examples/todo-notes/
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{hyperref}
\hypersetup{colorlinks=true,
linkcolor=[rgb]{.67 .27 .27},
citecolor=[rgb]{.09 .29 .54},
urlcolor=[rgb]{.09 .29 .54}}

% Set autoref text
% http://tex.stackexchange.com/a/36576/6784
\renewcommand*{\figureautorefname}{Fig.}
\renewcommand*{\equationautorefname}{Eq.}
\renewcommand*{\tableautorefname}{Table}
\renewcommand*{\sectionautorefname}{Section}

@@

% Add Figure, Table prefixes to references
% http://tex.stackexchange.com/a/6063/6784
\let\ref\autoref

@@

% \section{Todo}
% \listoftodos

@@

\section{Stability analysis of biological networks}\label{sec:stabanalbn}
\todo{Describe dynamical systems in terms of vector fields on Manifolds \cite{Deville}}
Consider an n-dimensional vector
$$
x(t) = (x_1(t), \ldots x_n(t)) = \vec{x}(t)
$$
whose components are solutions to the arbitrary first order system
$$
\frac{dx_i(t)}{dt} = F_i(\vec{x}(t)), \; (i=1,\ldots,n)
$$
where $F_i$ represent, potentially nonlinear, functions of the given vector of state variables. We can associate a directed graph $G_X$ that describes the manner in which each of the variables depends upon one another given by an adjacency matrix $\adj(G_X)$ where
 \begin{displaymath}
   \adj(G_X)_{ij} = \left\{
     \begin{array}{ll}
       1, & F_i \hbox{ depends on } x_j\\
       0, & F_i \hbox{ does not depend on } x_j
     \end{array}
   \right.
\end{displaymath}
For example, given the system $X = \{\{x_1,x_2\},\{F_1,F_2\}\}$
\begin{align*}
\frac{dx_1}{dt} & = F_1(x_2)\\
\frac{dx_2}{dt} & = F_2(x_1,x_2)
\end{align*}
we have the graph
\begin{center}
\input{tex/examplehamdyngraph}
\end{center}
with adjacency matrix
$$
\adj(G_X) = \begin{pmatrix}
0 & 1 \\
1 & 1
\end{pmatrix}
$$
\todo{Define strongly connected components and connect to operads \cite{Vagner2014}}

@@

The spectral abscissa of the matrix $A$ is defined as
$$
\eta(A) = \max_i \{\Re(\lambda_i)\}
$$
where $\lambda_i$ are the eigenvalues of $A$. The system defined by $F_i$ and $\vec{x}^0$ is dynamically stable if the spectral abscissa of $A$ is less than zero, equivalently, $\eta(A) < 0$. This is because the general solution to \ref{eq:lineardynsys} is
$$
y_i(t) = \sum_j b_{ij} e^{\lambda_j t}, \; (i=1,\ldots,n)
$$
for some matrix $B=(b_{ij})$ and thus all $\vec{y} = \vec{x} - \vec{x}^0$ decay to zero when all $\lambda_i < 0$.

@@

\section{Stability criteria}
Suppose that $M$ is an $n \times n$ matrix with real coefficients. Motivated by the discussion of \ref{sec:stabanalbn}, we will call the matrix $M$ stable if all its eigenvalues have negative real part.  Define the characteristic polynomial as $\chi(M)(z) = \det(zI - M)$. Denote the coefficients of $\chi(M)$ as $s_j$ and its roots as $r_j$, writing $p(x) = x^n + \sum_{k=1}^n s_k x^{n-k} = \prod_{k=1}^n (x-r_k)$.  Then, since the roots of $\chi$ are the eigenvalues of $M$, asking that $M$ be stable is equivalent to asking that the roots of $\chi$ all have negative real part.  We will formulate conditions for this to happen in terms of the coefficients of $\chi$.

@@

The simplest case is $n=2$.  There we have the regions $R_{2000},
R_{1100}, R_{0200}, R_{0020}, R_{0002}$ and our invariants are $D =
s_2, \delta = s_1^2 - 4s_2, \Delta = s_1$.  The relation between these
is summarized by the conditions
\begin{align}
R_{2000} \qquad &D > 0, \delta > 0, \Delta > 0, \cr
R_{1100} \qquad &D < 0, \delta > 0, \cr
R_{0200} \qquad &D > 0, \delta > 0, \Delta < 0, \cr
R_{0020} \qquad &\delta < 0, \Delta > 0, \cr
R_{0002} \qquad &\delta < 0, \Delta < 0. \cr
\end{align}
These regions used for the classification of stability of two-component systems are show in \ref{fig:region2x2}. The stable regions are given by $R_{2000}$ and $R_{0020}$. All of the other regions represent unstable matrices.

@@

Moving to $n=3$, our invariants now look as follows: $D = s_3, \delta
= 18 s_1 s_2 s_2 + s_1^2 s_2^2 - 4 s_2^3 - 4 s_1^3 s_3 - 27 s_3,
\Delta = s_1 s_2 - s_3$.  However, now we can no longer simply use their
signs to distinguish regions.  For instance, consider the regions
$R_{3000}$ and $R_{1200}$.  These both have $D > 0$ and $\delta > 0$
but to tell them apart, we need to note that the region $\delta > 0$
has two connected components and use the auxiliary condition that $s_1 > 0$ to
specify which of the components contains our region. The regions analogous to \ref{fig:region2x2} for three component systems where $D=1$ is given in \ref{fig:region3x3}.
%(Looking at the sign of $\Delta$ will not help here because, while $\Delta > 0$ in the region $R_{3000}$, it is also positive for some elements of $R_{1200}$ as well.)

@@

The probability of \emph{a priori} stability and of stability to perturbations for all three component systems is given in Table \ref{tab:structstabmat3}. \ref{fig:connectcycle3D3x3} shows the probability of stability to perturbations for all three component systems as a function of connectivity and number of cycles. \ref{fig:stab3x3} shows the probability of stability to perturbations for all three component systems as a function of connectivity. \ref{fig:cycle3x3} shows the probability of stability to perturbations for all three component systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}.

@@

% \input{tex/tablewithgraphs}
